---
title: "Group Dynamics Survival Analysis and Data visualization *(Karadimitriou et al In Prep)*"
author: "Lili Vizer and Natalia Karadimitriou"
date: "*Last run on `r format(Sys.time(), '%d %B %Y')`*"
output:
  html_document: 
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
# ## Knit standards
# knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, options(knitr.kable.NA = ''))
# options(warn = -1)

```


```{r load packages, message=FALSE, warning=FALSE, include=FALSE}
library(dplyr)
library(plyr)
library(tidyr)
library(curl)
library(survival)
library(survminer)
library(ggplot2)
library(ggfortify)
library(stringr)
library(kableExtra)
```

---

<br/>
This file contains all data manipulations, analysis and visualization associated with the probability of survival, settlement and recruitment of juvenile *A. percula* fish in adult tanks with varying anemone area.

We explore that:
i) does anemone size affect chances of survival in juveniles
ii) given that a juvenile survived, does anemone size affect its chance of settling
iii) given that a juvenile settled, does anemone size affect its chance of recruiting

this means that for every stage of the recruitment process we subset our dataset to only include the juveniles that have "passed" the previous stage

```{r color scheme, echo=FALSE, message=FALSE, warning=FALSE}
# "#e5f5f9" => survival
# "#99d8c9" => settlement
# "#2ca25f" => recruitment
# 
# 
# "#8c510a" => L_anemone
# "#d8b365" => M_anemone
# "#f6e8c3" => S_anemone
```


--
Same pipeline for every stage:
1. Kaplan-Meier curve        (survfit)   -> figure + median times
2. Log-rank test             (survdiff)  -> do curves differ?
3. Cox proportional hazards  (coxph)     -> hazard ratios between categories
4. PH assumption check       (cox.zph)   -> validity of the Cox model

--



## Survival Analysis {.tabset}

Kaplan-Meier curve (survfit) — descriptive, for the figure and median survival times.
Cox proportional-hazards regression (coxph) — gives the hazard ratios between anemone categories.
Log-rank test (survdiff) — tests whether the three survival curves differ overall.

```{r reading in survival dataset}
all_dataset <- read.csv("../Data/Karadimitriou_Surv_Set_Rec_Analysis_data_20260818.csv", header = T)
head(all_dataset)
colnames(all_dataset)
dim(all_dataset)
```

```{r removing rows not part of the experiment}
# 2 rows which are not in the experiment A31 & A34
all_data <- subset(all_dataset, !housing_tank_ID == "A31" & !housing_tank_ID == "A34")
nrow(all_data)
head(all_data)
```


```{r}
#table of fish survival, settlement and recruitment events by anemone category
data.frame(
  outcome = c("Survived", "Settled", "Recruited"),
  n = c(
    sum(all_data$survival_event    == 0, na.rm = TRUE),
    sum(all_data$settlement_event  == 1, na.rm = TRUE),
    sum(all_data$recruitment_event == 1, na.rm = TRUE)
  )
)

aggregate(
  cbind(
    Survived  = survival_event    == 0,
    Settled   = settlement_event  == 1,
    Recruited = recruitment_event == 1
  ) ~ Anemone_category,
  data = all_data,
  FUN = function(x) sum(x, na.rm = TRUE),
  na.action = na.pass
)

```




```{r survival result dataframe setup}
surv_data <- subset(all_data, select = c("housing_tank_ID", "fish_ID", "Anemone_category", "exp_day_survival", "survival_event"))
head(surv_data)
surv_data$Anemone_category <- as.factor(surv_data$Anemone_category) #needing to make anemone category a factor
```


```{r}
# --- Kaplan-Meier ---
probsurv <- Surv(time = surv_data$exp_day_survival, event = surv_data$survival_event)
KMsurv <- survfit(probsurv ~ Anemone_category, data = surv_data, type = "kaplan-meier")
KMsurv  # median survival times

# --- Log-rank: do the curves differ among categories? ---
surv_diff <- survdiff(Surv(exp_day_survival, survival_event) ~ Anemone_category, data = surv_data)
surv_diff

# --- Cox proportional hazards: hazard ratios ---
surv_cox <- coxph(Surv(exp_day_survival, survival_event) ~ Anemone_category, data = surv_data)
summary(surv_cox)

kable(as.data.frame(summary(surv_cox)$coef), digits = 3) %>%
  kable_styling(position = "center", font_size = 14, full_width = FALSE) %>%
  column_spec(1, bold = TRUE) %>%
  kable_classic(full_width = FALSE, html_font = "Times")

# --- Check the proportional-hazards assumption ---
cox.zph(surv_cox)
```


```{r survival figure, message=FALSE, warning=FALSE, fig.align='center'}
gg_surv <- ggsurvplot(fit = KMsurv,
  data = surv_data,
  conf.int = TRUE,
  conf.int.alpha = 0.09,      # much lighter ribbons so lines show through
  pval = TRUE,
  pval.size = 4,
  pval.coord = c(170, 0),        # move p-value out of the busy lower-left
  risk.table = FALSE,
  legend.labs = c("Large", "Medium", "Small"),  # spell out; clearer than L/M/S
  legend = "right",          # move legend to open upper-right space
  legend.title = "Anemone size",
  break.time.by = 25,
  censor.shape = "|",
  censor.size = 3,
  size = 1.1,                      # slightly thicker step lines
  xlab = "Time since introduction (days)",
  ylab = "Probability of survival",
  palette = c("#3b250e", "#8c510a", "#de913f"),
  ggtheme = theme_classic() +
    theme(legend.text = element_text(size = 11),
          legend.title = element_text(size = 12),
          axis.text = element_text(size = 11),
          axis.title = element_text(size = 12),
          legend.box.spacing = unit(1, "pt"),           # gap between panel and legend
          legend.margin = margin(0, 0, 0, 0),           # trim padding around legend
          legend.key.width = unit(0.5, "cm")))

gg_surv


ggsave("../Figures/Group_Dynamics_Survival_20260818.png", width = 6, height = 5, bg="white", dpi=1200) # save as PNG
ggsave("../Figures/Group_Dynamics_Survival_20260818.svg", width = 6, height = 5, bg="white", dpi=1200) # save as SVG
```

**Figure X**: The probability of *A. percula* juveniles surviving in adult tanks with S, M, or L anemone area. 

n: total number of subjects in each curve.
time: the time points on the curve.
n.risk: the number of subjects at risk at time t
n.event: the number of events that occurred at time t.
n.censor: the number of censored subjects, who exit the risk set, without an event, at time t.
lower,upper: lower and upper confidence limits for the curve, respectively.
strata: indicates stratification of curve estimation. If strata is not NULL, there are multiple curves in the result. The levels of strata (a factor) are the labels for the curves.


<br/>

## Settlement Analysis {.tabset}

0=not settled
1=settled


Kaplan-Meier curve plotted as cumulative incidence (fun = "event").
Logistic regression (glm, binomial) — odds of settlement by anemone category, with S set as the reference level.

```{r settlement dataframe setup}
#SETTLEMENT = only juveniles that SURVIVED (n = 27)
# survival_event == 0 means censored = persisted to end = survived
settle_data <- subset(all_data, survival_event == 0,
  select = c(housing_tank_ID, fish_ID, Anemone_category,
             exp_day_settlement, settlement_event))

settle_data$Anemone_category <- as.factor(settle_data$Anemone_category) #needing to make anemone category a factor
str(settle_data)

table(settle_data$Anemone_category, settle_data$settlement_event)
```


```{r}
# --- Kaplan-Meier ---
probsettle <- Surv(time = settle_data$exp_day_settlement, event = settle_data$settlement_event)
KMsettle <- survfit(probsettle ~ Anemone_category, data = settle_data, type = "kaplan-meier")
KMsettle

# --- Log-rank ---
settle_diff <- survdiff(Surv(exp_day_settlement, settlement_event) ~ Anemone_category, data = settle_data)
settle_diff

# --- Cox proportional hazards ---
settle_cox <- coxph(Surv(exp_day_settlement, settlement_event) ~ Anemone_category, data = settle_data)
summary(settle_cox)

kable(as.data.frame(summary(settle_cox)$coef), digits = 3) %>%
  kable_styling(position = "center", font_size = 14, full_width = FALSE) %>%
  column_spec(1, bold = TRUE) %>%
  kable_classic(full_width = FALSE, html_font = "Times")

# --- PH check ---
cox.zph(settle_cox)

# --- Separation check (L & M all settled -> Cox HRs unstable here) ---
table(settle_data$Anemone_category, settle_data$settlement_event)
# If separated, report the log-rank test + proportions above, not the Cox HR.
# fisher.test(table(settle_data$Anemone_category, settle_data$settlement_event))
```




```{r settlement figure, message=FALSE, warning=FALSE, fig.align='center'}
gg_settle <- ggsurvplot(fit = KMsettle,
  data = settle_data,
  conf.int = TRUE,
  conf.int.alpha = 0.09,      # much lighter ribbons so lines show through
  pval = TRUE,
  pval.size = 4,
  pval.coord = c(165, 0.75),       
  risk.table = FALSE,
  fun = "event",
  legend.labs = c("Large", "Medium", "Small"),  # spell out; clearer than L/M/S
  legend = "right",          # move legend to open upper-right space
  legend.title = "Anemone size",
  break.time.by = 25,
  censor.shape = "|",
  censor.size = 3,
  size = 1.1,                      # slightly thicker step lines
  xlab = "Time since introduction (days)",
  ylab = "Probability of settlement",
  palette = c("#3b250e", "#8c510a", "#de913f"),
  ggtheme = theme_classic() +
    theme(legend.text = element_text(size = 11),
          legend.title = element_text(size = 12),
          axis.text = element_text(size = 11),
          axis.title = element_text(size = 12),
          legend.box.spacing = unit(1, "pt"),           # gap between panel and legend
          legend.margin = margin(0, 0, 0, 0),           # trim padding around legend
          legend.key.width = unit(0.5, "cm")
          ))

gg_settle


ggsave("../Figures/Group_Dynamics_Settlement_20260818_v2.png", width = 6, height = 5, bg="white", dpi=1200) # save as PNG
ggsave("../Figures/Group_Dynamics_Settlement_20260818_v2.svg", width = 6, height = 5, bg="white", dpi=1200) # save as SVG

```



## Recruitment Analysis {.tabset}

0=not recruited
1=recruited

Kaplan-Meier curve (fun = "event"), L and M only.
Log-rank test (survdiff).
Logistic regression (glm, binomial), L and M only.

```{r recruitment dataset setup}
#RECRUITMENT — only juveniles that SETTLED (n = 19)
recruit_data <- subset(all_data, settlement_event == 1,
  select = c(housing_tank_ID, fish_ID, Anemone_category,
             exp_day_recuitment, recruitment_event))

# Recruitment regression subset: drop S (only 1 settled there which is then recruited, this will
# skew data, so dropping it) -> n = 18
recruit_LM <- droplevels(subset(recruit_data, Anemone_category %in% c("L", "M")))
recruit_LM$Anemone_category <- factor(recruit_LM$Anemone_category, levels = c("L", "M")) #set L as reference level

str(recruit_LM)
```


```{r}

# --- Kaplan-Meier --- (note column spelling: exp_day_recuitment)
probrecruit <- Surv(time = recruit_LM$exp_day_recuitment, event = recruit_LM$recruitment_event)
KMrecruit <- survfit(probrecruit ~ Anemone_category, data = recruit_LM, type = "kaplan-meier")
KMrecruit

# --- Log-rank ---
recruit_diff <- survdiff(Surv(exp_day_recuitment, recruitment_event) ~ Anemone_category, data = recruit_LM)
recruit_diff

# --- Cox proportional hazards ---
recruit_cox <- coxph(Surv(exp_day_recuitment, recruitment_event) ~ Anemone_category, data = recruit_LM)
summary(recruit_cox)

kable(as.data.frame(summary(recruit_cox)$coef), digits = 3) %>%
  kable_styling(position = "center", font_size = 14, full_width = FALSE) %>%
  column_spec(1, bold = TRUE) %>%
  kable_classic(full_width = FALSE, html_font = "Times")

# --- PH check ---
cox.zph(recruit_cox)
```


```{r recruitment figure, message=FALSE, warning=FALSE, fig.align='center'}
gg_recruit <- ggsurvplot(fit = KMrecruit,
  data = recruit_LM,
  conf.int = TRUE,
  conf.int.alpha = 0.09,      # much lighter ribbons so lines show through
  pval = TRUE,
  pval.size = 4,
  pval.coord = c(180, 0),        # move p-value out of the busy lower-left
  risk.table = FALSE,
  fun = "event",
  legend.labs = c("Large", "Medium"),  # spell out; clearer than L/M/
  legend = "right",          # move legend to open upper-right space
  legend.title = "Anemone size",
  break.time.by = 25,
  censor.shape = "|",
  censor.size = 3,
  size = 1.1,                      # slightly thicker step lines
  xlab = "Time since introduction (days)",
  ylab = "Probability of recruitment",
  palette = c("#3b250e", "#8c510a"),
  ggtheme = theme_classic() +
    theme(legend.text = element_text(size = 11),
          legend.title = element_text(size = 12),
          axis.text = element_text(size = 11),
          axis.title = element_text(size = 12),
          legend.box.spacing = unit(1, "pt"),           # gap between panel and legend
          legend.margin = margin(0, 0, 0, 0),           # trim padding around legend
          legend.key.width = unit(0.5, "cm")))

gg_recruit


ggsave("../Figures/Group_Dynamics_Recruitment_20260818.png", width = 6, height = 5, bg="white", dpi=1200) # save as PNG
ggsave("../Figures/Group_Dynamics_Recruitment_20260818.svg", width = 6, height = 5, bg="white", dpi=1200) # save as SVG
```




